---
title: "Fix species names v1 and v2"
author: "Brunno F Oliveira & Bioshifts group"
date: "Last compiled on `r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
---

# setup

```{r setup, message = FALSE, warning = FALSE}

require(data.table)
library(knitr)
library(dplyr)
library(DT)
library(tidyverse)
library(readxl)
library(traitdataform)
library(parallel)
library(pbapply)
library(taxize)
library(taxizedb)
library(taxadb)
#library(rcol)
library(ggrepel)
library(bdc)
library(rgbif)
library(googledrive)

```

# Load functions
```{r}

source("clean_taxa_functions.R")

```


# Load range shift data
```{r v1mv2, warning=FALSE}
# Load databases

# Load v1
biov1 <- read.table("/Users/nikkimoore/Documents/bioshifts-traits/data-raw/Shifts2018_checkedtaxo.txt",
                    header = T,
                    encoding="latin1")

# v1
biov1$sp_name_v1 <- biov1$New_name
# create a shift ID
biov1$S_ID_v1 <- sapply(1:nrow(biov1), function(x) paste("S",x,sep = "_"))

# some manual fixes
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Pterourus garamas")] <- "Papilio garamas"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Lestes inequalis")] <- "Lestes inaequalis"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Aegypius occipitalis")] <- "Trigonoceps occipitalis"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Parisoma subcaeruleum")] <- "Curruca subcoerulea"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Fixsenia prunoides")] <- "Fixsenia pruni"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Fixsenia eximia")] <- "Satyrium eximium"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Turbo cinereus")] =  "Lunella cinerea"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Excalfactoria adansonii")] =  "Coturnix adansonii"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Mimulus nanus")] =  "Diplacus nanus"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Berberis pinnata")] =  "Mahonia pinnata"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Hypomecis atomaria")] =  "Phalaena atomaria"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Hypaurotis smaragdinus")] =  "Chrysozephyrus smaragdinus"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Callophrys frivaldszkyi")] =  "Ahlbergia frivaldszkyi"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Fixsenia spini")] =  "Satyrium spini"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Zeheba aureata")] =  "Zeheba aureatoides"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Tharsalea epixanthe")] =  "Lycaena epixanthe"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Nymphalis comma")] =  "Polygonia comma"
# biov1$sp_name_v1[which(biov1$sp_name_v1 == "Nymphalis interrogationis")] =  "Polygonia interrogationis"

# reported name = Publi_name
biov1$sp_reported_name_v1 = Clean_Names(gsub("_"," ",biov1$Publi_name), return_gen_sps = F)
# accepted name = New_name
biov1$sp_name_v1 = Clean_Names(gsub("_"," ",biov1$New_name), return_gen_sps = T)

# Load v2 
# biov2 <- read.csv("v2/Bioshifts.v2.final.csv")
biov2 <- read_excel("/Users/nikkimoore/Documents/bioshifts-traits/data-raw/Bioshifts.v2.final.corrected.xlsx", 
                    sheet = 2) # new file sent by Sarah with start/end dates for shifts
# create a shift ID
biov2$S_ID_v2 <- sapply(1:nrow(biov2), function(x) paste("S",x,sep = "_"))

# v2
# reported name = Scientific Name
biov2$sp_reported_name_v2 = Clean_Names(gsub("_"," ",biov2$`Scientific Name`), return_gen_sps = F)
which(biov2$sp_reported_name_v2=="")
# accepted name = species
biov2$sp_name_v2 = Clean_Names(gsub("_"," ",biov2$species), return_gen_sps = T)

# manual
biov2$sp_reported_name_v2[which(biov2$sp_reported_name_v2 == "Coprosma 3sp")] <- "Coprosma sp."
biov2$sp_name_v2[which(biov2$sp_name_v2 == "Coprosma 3sp")] <- "Coprosma sp."
# potential missing sp names
# test <- sapply(biov2$sp_name_v2, function(x){
#     length(strsplit(x," ")[[1]])
# })
# test2 <- sapply(biov2$sp_reported_name_v2, function(x){
#     length(strsplit(x," ")[[1]])
# })
# View(biov2[which(test == 1 & test2>1),])

```

# Fix species names

```{r eval=FALSE}

splist <- data.frame(reported_name = unique(c(biov1$sp_reported_name_v1, biov2$sp_reported_name_v2)))
splist$v1 <- 0
splist$v1[which(splist$reported_name %in% biov1$sp_reported_name_v1)] <- 1
splist$v2 <- 0
splist$v2[which(splist$reported_name %in% biov2$sp_reported_name_v2)] <- 1
splist$reported_name <- gsub("_"," ",splist$reported_name)

# add original accepted
original_accepted_v1 <- 
    data.frame(
        species = gsub("_"," ",biov1$sp_name_v1),
        reported_name = gsub("_"," ",biov1$sp_reported_name_v1))
original_accepted_v2 <- 
    data.frame(
        species = gsub("_"," ",biov2$sp_name_v2),
        reported_name = gsub("_"," ",biov2$sp_reported_name_v2))

original_accepted <- rbind(original_accepted_v1,original_accepted_v2)

splist <- merge(splist,original_accepted, by="reported_name", all.x = TRUE)
splist <- splist[-which(duplicated(paste(splist$reported_name,splist$species))),]

# fix issues with original name
# V2 never uses subspecies or variants, so look for species at the genus level
splist$reported_name_fixed <- Clean_Names(splist$reported_name,
                                          return_gen_sps = TRUE)

# do not look at species identified to the genus level or cf. species
spgen <- sapply(splist$reported_name, function(x){
    tmp. <- strsplit(x, " ")[[1]]
    any(tmp. == "sp" | tmp. == "spp" | tmp. == "sp." | tmp. == "spp." | tmp. == "cf." | tmp. == "cf" | length(tmp.) == 1)
})
spgen <- splist$reported_name[spgen]
all(spgen %in% splist$reported_name)

V3 <- data.frame(matrix(nrow = nrow(splist), ncol = 7))
names(V3) = c("scientificName", "kingdom", "phylum", "class", "order", "family", "db")
splist <- cbind(splist, V3)

# For these species, use classification from the original database
for(i in 1:length(spgen)){
    tofill <- which(splist$reported_name == spgen[i])
    db <- splist[tofill,2:3]
    db <- names(db)[which(db==1)]
    if(length(db) > 1) { db = "v1"}
    if(db == "v1"){
        v1row <- which(biov1$sp_reported_name_v1 == spgen[i])
        splist$scientificName[tofill] <- unique(biov1$sp_name_v1[v1row])[1]
        splist$kingdom[tofill] <- unique(biov1$Kingdom[v1row])[1]
        splist$phylum[tofill] <- unique(biov1$Phylum[v1row])[1]
        splist$class[tofill] <- unique(biov1$Class[v1row])[1]
        splist$order[tofill] <- unique(biov1$Order[v1row])[1]
        splist$family[tofill] <- unique(biov1$Family[v1row])[1]
        splist$db[tofill] <- "Bioshifts_v1"
    }
    if(db == "v2"){
        v2row <- which(biov2$sp_reported_name_v2 == spgen[i])
        splist$scientificName[tofill] <- unique(biov2$sp_name_v2[v2row])[1]
        splist$kingdom[tofill] <- unique(biov2$kingdom[v2row])[1]
        splist$phylum[tofill] <- unique(biov2$phylum[v2row])[1]
        splist$class[tofill] <- unique(biov2$class[v2row])[1]
        splist$order[tofill] <- unique(biov2$order[v2row])[1]
        splist$family[tofill] <- unique(biov2$family[v2row])[1]
        splist$db[tofill] <- "Bioshifts_v2"
    }
}

splist$db_code <- NA
splist$db_code <- as.character(splist$db_code)

splist <- splist[,c("species","v1", "v2", "scientificName", "kingdom", "phylum", "class", "order", "family", "db", "db_code", "reported_name_fixed", "reported_name")]

```

Look for accepted names

We use a combination of 'taxadb' and 'bdc' packages. taxadb builds on taxize, but it is much faster because it uses species lists from providers stored in the HD. The advantage of 'bdc' over taxadb is that it allows using fuzzy matching. As fuzzy matching can take longer times when running over a long list of species, we first run the function bdc_query_names_taxadb on exact matches, then use fuzzy matching on the remaining species.

## Species to find
```{r}

tofind_ <- splist[which(is.na(splist$scientificName)),]
tofind_ <- unique(c(tofind_$species,tofind_$reported_name_fixed))

tofind <- data.frame(matrix(nrow = length(tofind_), ncol = 8))
names(tofind) = c("scientificName", "kingdom", "phylum", "class", "order", "family", "db", "db_code")

tofind <- data.frame(species = tofind_, tofind)

tofind <- tofind %>%
    mutate(across(everything(), as.character))

```

## GBIF
```{r eval=FALSE}

togo <- tofind[which(is.na(tofind$scientificName)),]

mycols <- c("speciesKey", "kingdom","phylum","class","order","family", "species")

# retrieve sp names
cl <- makeCluster(detectCores()-2)
clusterExport(cl, c("togo","standardize_taxa"))

gbif_names <- pblapply(togo$species, function(x){
    try(standardize_taxa(data.frame(verbatimScientificName = x), 
                         fuzzy = FALSE,
                         silent = TRUE))
}, cl = cl)

stopCluster(cl)

rem <- sapply(gbif_names, class)
rem <- which(rem=="try-error")
gbif_names <- gbif_names[-rem]
gbif_names <- rbindlist(gbif_names)

gbif_names <- gbif_names[-which(is.na(gbif_names$scientificName)),]
gbif_names <- gbif_names[which(gbif_names$taxonRank=='species'),]

# remove duplicates
if(any(duplicated(gbif_names$verbatimScientificName))){
    gbif_names <- gbif_names[-which(duplicated(gbif_names$verbatimScientificName)),]
}

any(!gbif_names$original_search %in% splist$species)

cat("--- Summary ---\n",
    "N taxa:",nrow(togo),"\n",
    "N taxa found:",nrow(gbif_names), "\n",
    "N taxa not found:", nrow(togo)-nrow(gbif_names))

gbif_names <- gbif_names[,c("verbatimScientificName","scientificName","kingdom","phylum","class","order","family","taxonID")]
names(gbif_names) <- c("species","scientificName","kingdom","phylum","class","order","family","db_code")
gbif_names$db <- "gbif"
gbif_names$db_code <- gsub("http://www.gbif.org/species/","",gbif_names$db_code)
gbif_names$db_code <- paste("GBIF:",gbif_names$db_code,sep = "")

# Feed
tofind <- tofind %>% 
    rows_patch(gbif_names, 
               by = "species")

gc()

```

## ITIS
```{r eval=FALSE}

togo <- tofind[which(is.na(tofind$scientificName)),]

# retrieve sp names
itis_names <- Find_Names(spnames = togo$species,
                         db = "itis",
                         suggest_names = FALSE, # using exact names
                         return_accepted_only = TRUE) 

# remove duplicates
if(any(duplicated(itis_names$original_search))){
    itis_names <- itis_names[-which(duplicated(itis_names$original_search)),]
}

any(!itis_names$original_search %in% splist$species)

itis_names <- itis_names[,c("original_search","scientificName","kingdom","phylum","class","order","family","acceptedNameUsageID")]
names(itis_names) <- c("species","scientificName","kingdom","phylum","class","order","family","db_code")
itis_names$db <- "itis"

# Feed
tofind <- tofind %>% 
    rows_patch(itis_names, 
               by = "species")

gc()

```

## NCBI
```{r eval=FALSE}

togo <- tofind[which(is.na(tofind$scientificName)),]

# retrieve sp names
ncbi_names <- Find_Names(spnames = togo$species,
                         db = "ncbi",
                         suggest_names = FALSE, # using exact names
                         return_accepted_only = TRUE) 

# remove duplicates
if(any(duplicated(ncbi_names$original_search))){
    ncbi_names <- ncbi_names[-which(duplicated(ncbi_names$original_search)),]
}

any(!ncbi_names$original_search %in% splist$species)

ncbi_names <- ncbi_names[,c("original_search","scientificName","kingdom","phylum","class","order","family","acceptedNameUsageID")]
names(ncbi_names) <- c("species","scientificName","kingdom","phylum","class","order","family","db_code")
ncbi_names$db <- "ncbi"

# some names from ncbi come with duplicated genus
# remove duplicated names
new <- sapply(ncbi_names$scientificName, function(x){
    tmp <- strsplit(x," ")[[1]]
    if(any(duplicated(tmp))){
        paste(tmp[-1], collapse = " ")
    } else {
        x
    }
})

ncbi_names$scientificName <- new

# Feed
tofind <- tofind %>% 
    rows_patch(ncbi_names, 
               by = "species")

gc()

```


## IUCN
```{r eval=FALSE}

togo <- tofind[which(is.na(tofind$scientificName)),]

# retrieve sp names
## PS: Cant filter on iucn method
iucn_names<-bdc_query_names_taxadb(togo$species, 
                                   db = "iucn",
                                   suggest_names = FALSE) # using exact names
iucn_names$scientificName <- paste(iucn_names$genus,iucn_names$specificEpithet) 

# remove duplicates
if(any(duplicated(iucn_names$original_search))){
    iucn_names <- iucn_names[-which(duplicated(iucn_names$original_search)),]
}

any(!iucn_names$original_search %in% splist$species)

# select only accepted names
iucn_names <- iucn_names[which(iucn_names$taxonomicStatus == "accepted"),]

cat("--- Summary ---\n",
    "N taxa:",nrow(togo),"\n",
    "N taxa found:",length(which(iucn_names$taxonomicStatus == "accepted")), "\n",
    "N taxa not found:", nrow(togo)-length(which(iucn_names$taxonomicStatus == "accepted")))

iucn_names <- iucn_names[,c("original_search","scientificName","kingdom","phylum","class","order","family","acceptedNameUsageID")]
names(iucn_names) <- c("species","scientificName","kingdom","phylum","class","order","family","db_code")
iucn_names$db <- "iucn"

# Feed
tofind <- tofind %>% 
    rows_patch(iucn_names, 
               by = "species")

gc()

```


## Fix issues with scientific names
```{r}

# some names from ncbi come with duplicated genus
# remove duplicated names
# check <- sapply(tofind$scientificName, function(x){
#     tmp <- strsplit(x," ")[[1]]
#     any(duplicated(tmp))
# })
# 
# View(tofind[which(check),]) # all ncbi

new <- sapply(tofind$scientificName, function(x){
    tmp <- strsplit(x," ")[[1]]
    if(any(duplicated(tmp))){
        paste(tmp[-1], collapse = " ")
    } else {
        x
    }
})

tofind$scientificName <- new

```

## Add found species names to the splist
```{r}

all(tofind$species %in% splist$species | tofind$species %in% splist$reported_name_fixed)

found <- tofind[-which(is.na(tofind$scientificName)),]

for(i in 1:length(found$species)){
    tofill <- unique(c(which(splist$species == found$species[i]),which(splist$reported_name_fixed == found$species[i])))
    splist$scientificName[tofill] <- found$scientificName[i]
    splist$kingdom[tofill] <- found$kingdom[i]
    splist$phylum[tofill] <-found$phylum[i]
    splist$class[tofill] <- found$class[i]
    splist$order[tofill] <- found$order[i]
    splist$family[tofill] <- found$family[i]
    splist$db[tofill] <- found$db[i]
    splist$db_code[tofill] <- found$db_code[i]
}

```


## Keep original names for species we couldnt find at any database
```{r eval=FALSE}

splist_ <- splist
# splist <- splist_
spnotfound <- unique(splist$reported_name[which(is.na(splist$scientificName))])

for(i in 1:length(spnotfound)){ cat('\r',i)
    tofill <- which(splist$reported_name == spnotfound[i] & is.na(splist$db))
    db <- splist[tofill,2:3]
    db <- as.character(na.omit(names(db)[which(db==1)]))
    if(length(db) > 1) { db = "v1"}
    if(db == "v1"){
        v1row <- which(biov1$sp_reported_name_v1 == spnotfound[i])
        splist$scientificName[tofill] <- unique(biov1$sp_name_v1[v1row])[1]
        splist$kingdom[tofill] <- unique(biov1$Kingdom[v1row])[1]
        splist$phylum[tofill] <- unique(biov1$Phylum[v1row])[1]
        splist$class[tofill] <- unique(biov1$Class[v1row])[1]
        splist$order[tofill] <- unique(biov1$Order[v1row])[1]
        splist$family[tofill] <- unique(biov1$Family[v1row])[1]
        splist$db[tofill] <- "Bioshifts_v1"
    }
    if(db == "v2"){
        v2row <- which(biov2$sp_reported_name_v2 == spnotfound[i])
        splist$scientificName[tofill] <- unique(biov2$sp_name_v2[v2row])[1]
        splist$kingdom[tofill] <- unique(biov2$kingdom[v2row])[1]
        splist$phylum[tofill] <- unique(biov2$phylum[v2row])[1]
        splist$class[tofill] <- unique(biov2$class[v2row])[1]
        splist$order[tofill] <- unique(biov2$order[v2row])[1]
        splist$family[tofill] <- unique(biov2$family[v2row])[1]
        splist$db[tofill] <- "Bioshifts_v2"
    }
}

all(tofind$species %in% splist$species | tofind$species %in% splist$reported_name_fixed)

```

## GBIF priority
Query using accepted names found on other databases
```{r eval=FALSE}

togo <- tofind[which(!tofind$db=="gbif"),]

# retrieve sp names
cl <- makeCluster(detectCores()-2)
clusterExport(cl, c("togo","standardize_taxa"))

gbif_names <- pblapply(togo$species, function(x){
    try(standardize_taxa(data.frame(verbatimScientificName = x),
                         silent = TRUE))
}, cl = cl)

stopCluster(cl)

rem <- sapply(gbif_names, class)
rem <- which(rem=="try-error")
gbif_names <- gbif_names[-rem]
gbif_names <- rbindlist(gbif_names)

gbif_names <- gbif_names[-which(is.na(gbif_names$scientificName)),]
gbif_names <- gbif_names[which(gbif_names$taxonRank=='species'),]

# remove duplicates
if(any(duplicated(gbif_names$verbatimScientificName))){
    gbif_names <- gbif_names[-which(duplicated(gbif_names$verbatimScientificName)),]
}

cat("--- Summary ---\n",
    "N taxa:",nrow(togo),"\n",
    "N taxa found:",nrow(gbif_names), "\n",
    "N taxa not found:", nrow(togo)-nrow(gbif_names))

if(any(gbif_names$scientificName)){
    gbif_names <- gbif_names[,c("verbatimScientificName","scientificName","kingdom","phylum","class","order","family","taxonID")]
    names(gbif_names) <- c("species","scientificName","kingdom","phylum","class","order","family","db_code")
    gbif_names$db <- "gbif"
    gbif_names$db_code <- gsub("http://www.gbif.org/species/","",gbif_names$db_code)
    gbif_names$db_code <- paste("GBIF:",gbif_names$db_code,sep = "")
    
    pos <- match(gbif_names$species,splist$scientificName,nomatch = NULL)
    
    splist$scientificName[pos] <- gbif_names$scientificName
    splist$kingdom[pos] <- gbif_names$kingdom
    splist$phylum[pos] <- gbif_names$phylum
    splist$class[pos] <- gbif_names$class
    splist$order[pos] <- gbif_names$order
    splist$family[pos] <- gbif_names$family
    splist$db[pos] <- gbif_names$db
    splist$db_code[pos] <- gbif_names$db_code
}

gc()

```

# save species list
```{r eval=FALSE}

splist$scientificName <- gsub("  "," ",splist$scientificName)

write.csv(splist, "output/splist.csv", row.names = F)

# species = accepted name from the DB (Genus+species) removing special characters 
# reported_name = species name as published in the original paper extracted from the DBs
# reported_name_fixed = original name removing special characters 
# scientificName is the accepted name following taxonomy harmonization applied above. For species we didn't find a name, we used the accepted name as reported in the DBs.

# upload to google drive
drive_upload("output/splist.csv",
             "BIOSHIFTS Working Group/DATA/splist_v3.csv")

```




# Fix
```{r}

splist <- read.csv("output/splist.csv")

splist$id_temp <- paste(splist$reported_name,splist$species)

# Fix
# V1
biov1$id_temp <- paste(biov1$sp_reported_name_v1,biov1$sp_name_v1)

biov1 <- merge(biov1, 
               splist[,c("scientificName", "reported_name_fixed","id_temp")], 
               by = "id_temp", 
               all.x = TRUE)
# View(biov1[,c("sp_reported_name_v1", "scientificName", "id_temp")])

biov1 <- biov1[,-which(colnames(biov1) == "id_temp")]

names(biov1)[which(names(biov1) == "scientificName")] <- "sp_name_std_v1"
biov1$sp_reported_name_v1 <- biov1$reported_name_fixed
biov1 <- biov1[-which(names(biov1) %in% c("reported_name_fixed","species"))]

# V2
biov2$id_temp <- paste(biov2$sp_reported_name_v2,biov2$sp_name_v2)
biov2 <- merge(biov2, 
               splist[,c("scientificName", "reported_name_fixed","id_temp")], 
               by = "id_temp", 
               all.x = TRUE)
# View(biov2[,c("sp_reported_name_v2", "scientificName", "id_temp")])
biov2 <- biov2[,-which(colnames(biov2) == "id_temp")]

names(biov2)[which(names(biov2) == "scientificName")] <- "sp_name_std_v2"
biov2$sp_reported_name_v2 <- biov2$reported_name_fixed
biov2 <- biov2[-which(names(biov2) %in% c("reported_name_fixed","species"))]


###################
# last step
biov1$sp_name_v1 <- gsub(" ","_",biov1$sp_name_v1)
biov1$sp_name_std_v1 <- gsub(" ","_",biov1$sp_name_std_v1)
biov1$sp_reported_name_v1 <- gsub(" ","_",biov1$sp_reported_name_v1)

biov2$sp_name_v2 <- gsub(" ","_",biov2$sp_name_v2)
biov2$sp_name_std_v2 <- gsub(" ","_",biov2$sp_name_std_v2)
biov2$sp_reported_name_v2 <- gsub(" ","_",biov2$sp_reported_name_v2)

###################
# sp_name_std = harmonized species name
# sp_name_v1 = accepted species name extracted from DBs removing special characters 
# sp_reported_name_v1 = species name reported from original papers, extracted from DBs, removing special characters 

```

# save DBs
```{r eval=FALSE}

write.csv(biov1, "v1/biov1_fixednames.csv", row.names = F)
write.csv(biov2, "v2/biov2_fixednames.csv", row.names = F)

```




***